Storm Data Analysis - Week 4

### 2021-03-20 Manuscript.Rmd
First round: data prep and changes to economical variables

suppressMessages(library(tidyverse))
```

@@

```{r data-prep}

@@

# Get the data
if (!file.exists("data/stormdata.csv")) {
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "data/stormdata.csv")
}

data <- read_csv("data/stormdata.csv")

@@

>>>>

suppressMessages(library(tidyverse))
# library(skimr)
```

@@

```{r data-prep, message=FALSE, warning=FALSE, cache = TRUE}

@@

# Get the data
if (!file.exists("data/stormdata.csv.bz2")) {
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "data/stormdata.csv.bz2")
}

# Read data in. fread seems way faster than other solutions.
data <- data.table::fread("data/stormdata.csv.bz2", data.table = FALSE)

@@

We create 2 new table columns; one is a properly-formatted `POSIXct` column including date and time information from columns `BGN_DATE`and `BGN_TIME`. A second column will be created in order to reconstruct information on property damage: columns `PROPDMGEXP` specifies an "exponent" for `PROPDMG`, in the form of a mathematical suffix: K for thousands, M for millions, and so on... We will compute a column containing the actual amount as an integer or double. Same operation will be performed for `CROPDMG` and `CROPDMGEXP`.

```{r column-formatting}

# Formatted data
fdata <- data %>%
    mutate(
        # Paste date and time For some reason, data includes a fixed time 00:00
        # which we don't need, so we split and get the first fragment
        temp_datetime = paste(strsplit(BGN_DATE, " ")[[1]][1], BGN_TIME),
        # Convert to datetime; 
        utc_datetime = strptime(
            temp_datetime,
            "%m/%d/%Y %H%M"),
        # correct timezones (there are older/unsupported abbreviations)
        c_timezone = recode(TIME_ZONE,
                            CDT = "US/Central",
                            CST = "US/Central",
                            UNK = "UTC",
                            MDT = "US/Mountain",
                            AST = "US/Eastern",
                            ADT = "US/Eastern",
                            CSt = "US/Central",
                            CSC = "UTC",
                            SCT = "US/Central",
                            ESY = "US/Eastern",
                            SST = "US/Samoa",
                            AKS = "US/Alaska",
                            GST = "US/Central"
                            ),
        # # Convert each time to its own time zone
        datetime = lubridate::force_tzs(utc_datetime, tzones = c_timezone)
    )
```

`*EXP` values are actually multipliers for variables indicating damages. The full legend can be found [here](ane.st/HandleExponent6c0f3 "Legend for EXP values").

```{r exponents}
unique(data$PROPDMGEXP)
unique(data$CROPDMGEXP)
```

We will transform them into the actual multiplier so we can compute actual damages. It is known that there are a few malformed values [*e.g.*: empty strings ("")]; they will be set to 0. According to the above analysis, malformed values are negligible both in numbers and absolute values.

```{r exp-to-multipliers}
fdata <- fdata %>%
    mutate(
        prop_mult = case_when(
            PROPDMGEXP %in% c(0, 1, 2, 3, 4, 5, 6, 7, 8) ~ 10,
            PROPDMGEXP %in% c('m', 'M') ~ 1000000,
            PROPDMGEXP %in% c('k', 'K') ~ 1000,
            PROPDMGEXP %in% c('b', 'B') ~ 1000000000,
            PROPDMGEXP %in% c('h', 'H') ~ 100,
            PROPDMGEXP == '+' ~ 1,
            TRUE ~ 0
        ),
        crop_mult = case_when(
            PROPDMGEXP %in% c(0, 1, 2, 3, 4, 5, 6, 7, 8) ~ 10,
            CROPDMGEXP %in% c('m', 'M') ~ 1000000,
            CROPDMGEXP %in% c('k', 'K') ~ 1000,
            CROPDMGEXP %in% c('b', 'B') ~ 1000000000,
            CROPDMGEXP %in% c('h', 'H') ~ 100,
            CROPDMGEXP == '+' ~ 1,
            TRUE ~ 0
        ),
        prop_dmg = PROPDMG * prop_mult,
        crop_dmg = CROPDMG * crop_mult
    )

We can get the number of invalid `EXP` values resulting in `0`-value multipliers using the following code:

```{r invalid-exps}
# Property damage
inv_prop_exp <- filter(fdata, PROPDMG != 0, prop_mult == 0) %>%
    select(PROPDMG, PROPDMGEXP, prop_mult, BGN_DATE) %>%
    nrow()

inv_crop_exp <- filter(fdata, CROPDMG != 0, crop_mult == 0) %>%
    select(CROPDMG, CROPDMGEXP, crop_mult, BGN_DATE) %>%
    nrow()
```

Using **inline code evaluation**, we can report `r inv_prop_exp` invalid property damage exponents, and `r inv_crop_exp` invalid crop damage exponents.
